Online Sequential Monte Carlo smoother for partially observed diffusion processes

### 2017-09-14 TexSubmission/answer_referees.tex

The importance of these quantities is now clearly stated after equation (2). We added the usual applications to the computation of the Fisher score and the intermediate quantity of the EM algorithm. We also highlighted that it was crucial for online inference of latent data models which supports the use of the proposed online smoother.

\item {\em General practical strategies to specify $\vartheta_k$, $p_k$, $\hat{\sigma}_+^k$ in the SDE context should be
discussed.}

\vspace{.3cm}

Choices that can be made for  $\vartheta_k$ and $p_{k-1}$ are discussed after (4) and the definition of $\phi_k^N$. The specific approximation of the fully adapted particle filter is also illustrated in the numerical section. The $\widehat{\sigma}_+^k$ is an upper bound that must be computed depending on which of the assumptions (A1)-(A3) is satisfied, its choice  is not in this sens completely left to the  user. However, as discussed at the end of page 8, it can be better, even if (A1) is satisfied, to choose the bound defined by (A3). The first one would (in general) require numerical optimization and provide a general bound (potentially far from the one obtained with the set of particles). The other would require $N^2$ computations ($N$ being the number of particles), but provides the optimal bound for the acceptance reject algorithm of Lemma 1. In addition to the existing paragraph, this is more explicitely discussed now in the numerical section, specially for the SINE model.

@@

\item {\em Conditions (A1)-(A3) should be carefully discussed. How restrictive are they? For
example, what if $\phi(\omega_s)$ (or $L_\omega$) is not bounded below? Is there any general solution
in this case?}

\vspace{.3cm}

Assumptions (A1)-(A3) are discussed at the end of Section~3. As soon as $\phi(\omega_s)$ (or $L_\omega$) is bounded below, (A3) is satisfied  for instance with the GPE based on the  EA3 procedure  which allows to use our algorithm with GPEs to approximate the transition densities. When $\phi(\omega_s)$ is not bounded below, this may indicate that alternative to GPEs should be used to approximate the transition densities. We suggest in Section~2 the use of continuous importance sampling estimators which require other sets of assumptions. We did not comment assumptions (A1)-(A3) for these methods as we focus on GPEs in the paper. It is important to note that we stress that any unbiased estimator is a candidate to be used in our algorithm and that the way assumptions (A1)-(A3) are checked depends on the specific choice made by the user.

@@

\item {\em The authors should discuss how to obtain $L_w$ and $U_w$, and the efficiency of their
algorithm in terms of these bounds. In particular, they should reference the layered
BB construction - Beskos et al. (2008). Also, they can take advantage of the factorisation (into two terms) in (9) to perform the reject-accept step of the algorithm
in Lemma 1 using two coins - note that the first term does not depend on BB points
so a fail in the first coin would make the simulation of the BB points unnecessary.}

\vspace{.3cm}

The performance of the estimator of the transition density depends on the choice of $\Lo_\mw$ and $\U_\mw$ which is specific to the SDE. In the case of the models analyzed in the numerical section these bounds are discussed in [13] for the SINE model and in [27] for the log-growth model. We added a remark to stress the fact that  in the case where $\phi$ is not upper bounded, [5] proposed the EA3 algorithm. This {\em layered Brownian bridge} construction first samples random variables to determine in which {\em layer} the Brownian bridge lies before simulating the bridge conditional on the event that it belongs to the layer. By continuity of $\phi$, $\Lo_\mw$ and  $\U_\mw$ can be computed easily.

We agree with the Referee that the specific form of the GPE estimate allows to improve (slightly) its computational cost. Nonetheless, as discussed after Lemma~1 and in the numerical section, the cost of the AR algorithm depends mainly on the choice of the upper bound for $\widehat{q}_k$ (see answer to item 7. with numerical experiments).

@@

This is now mentioned at the end of Section~3.

\item {\em Concerning the paragraph starting in page 8 and ending in page 9. Can the authors
provide some analytical result about when that strategy is better as a function of the
rejection sampling global accep. prob. and the cost to sample (layered) Brownian
bridges?}

\vspace{.3cm}

This important problem is model dependent. Indeed, if assumption $(A1)$ is satisfied, to perform the rejection sampling, the user has to either
\begin{enumerate}
\item find $\widehat{\sigma}_+^k$ by numerical optimization in the (A1) and (A2) case, and set for every particle $i$, the upper bound to $\widehat{\sigma}_+^k$ ;
\item or, for each particle $i$, find a $\widehat{\sigma}_+^{k, i } = \mathrm{sup}_{j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^i_{k+1},\zeta)$, in the (A3) case.
\end{enumerate}
For each $i$, the corresponding bound is then used to apply Lemma 1. The first bound is uniform and does not depend on the particles, therefore can be used for every particle $i$.  However, the resulting bound  can be "large"(with respect to the simulated set of particles) for the AR algorithm of Lemma 1.
The second bound requires $N$ computations per particle (therefore $N^2$ computations). However, it is clear that this second bound gives the optimal bound for Lemma 1.

Actually, the first bound is more likely to be "too large" when the time step is large. Furthermore, in that case, a rejection in Lemma 1 is more costly as it would require sampling another Brownian bridge skeleton (which would be large due to the large time step). Overall, it seems difficult to provide an analytical result about which upper bound to choose. However, we did an empirical study on the SINE model, that satisfies A1 and A3. Results are presented in Figure \ref{Figure}.

 As expected from what is said above, for a large time step, even for a large number of particles, the bound given by A3, despite its quadratic cost, leads to less computational time than the uniform bound A1. 
 On the other hand, for a small time step, the uniform bound of A1 is good enough, and it's better to choose this bound instead of the optimal bound A3.
\begin{figure}[H]
\centering
\begin{tabular}{cc}
$\Delta = 0.5$ & $\Delta = 2$\\
\includegraphics[width = 0.4\textwidth]{Times_Delta05}&
\includegraphics[width = 0.4\textwidth]{Times_Delta2}
\end{tabular}
\caption{Empirical distribution of required time (in logarithmic scale of milliseconds) for the acceptation of a candidate in the acceptance rejection algorithm of Lemma 1, depending on which bound is chosen. A1 is when the uniform bound is chosen and A3 is when the particle based bound is chosen. The best choice depends on the number of particles (as expected) but also on the time step.}\label{Figure}
\end{figure}
This is now mentioned (without this empirical study) in the numerical experiments section (in the SINE model subsection).

@@

This was changed as suggested.

\item {\em page 11, line 10: does the fixed lag algorithm have the same cost for all lag values?
What is the computational time of one run of GRand PaRIS for the examples presented?}

\vspace{.3cm}

The only cost difference for two lags $\delta_1 < \delta_2$ is the memory space to keep tracks of particles between times $t+\delta_1$ and $t+\delta_2$ (basically, a $N*(\delta_2 - \delta_1)$ matrix). Therefore the computational time is identical for two different lags. On a personal laptop  (i7-6600U CPU @ 2.60GHz), for the parameters given in the paper, it took around 25 seconds for each E step. As mentioned in the paper, all the comparisons are made for a same computational time (that is why there is more particles used in the fixed lag techniques). 
This is now clarified in the paper.

### 2017-09-14 TexSubmission/ParisEM.tex
Last modifications

\begin{abstract}
This paper introduces a new algorithm to approximate smoothed additive functionals for partially diffusion processes. 
This method relies on a new Sequential Monte Carlo method which allows to compute such approximations online, i.e. as the observations are received, and with a computational complexity growing linearly with the number of Monte Carlo samples. The original algorithm cannot be used in the case of partially observed stochastic differential equations since the transition density of the latent data is usually unknown. 
We prove that it may be extended to partially observed continuous processes by replacing this unknown quantity by an unbiased estimator obtained for instance using general Poisson estimators. 
This estimator is consistent and its performance are illustrated using data from two models. 
\end{abstract}

@@

\section{Introduction}
This paper introduces a new algorithm to solve the smoothing problem for hidden Markov models (HMMs) whose hidden state is a solution to a stochastic differential equation (SDE). In \cite{olsson:strojby:2011}, these models are referred to as partially observed diffusion (POD) processes. The hidden state process $(X_t)_{t\ge 0}$ is assumed to be a solution to a SDE and the only information available is given by noisy observations $(Y_{k})_{0\le k\le n}$ of the states $(X_k)_{0\le k\le  n}$ at some discrete time points $(t_k)_{0\le k\le n}$. The bivariate stochastic process $\{(X_{k},Y_{k})\}_{0\le k\le n}$ is a state space model such that conditional on the state sequence $(X_{k})_{0\le k\le n}$ the observations $(Y_{k})_{0\le k\le n}$ are independent and for all $0\le \ell\le n$ the conditional distribution of $Y_{\ell}$ given $\{X_{k}\}_{0\le k\le n}$ depends on $X_{\ell}$ only.

@@

However, in most cases the exact computation of these expectations is usually not possible explicitly. Sequential Monte Carlo (SMC) methods are popular algorithms to approximate smoothing distributions with random particles associated with importance weights. 
 \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with resampling steps to duplicate or discard particles according to their importance weights.
In the case of HMMs, approximations of the smoothing distributions may be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. 
Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. 
Then, a backward pass is performed to compute new weights (FFBS) or sample trajectories (FFBSi) in order to approximate the smoothing distributions. Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.\ using the observations as they are received) smoothed expectations of additive functionals. 
Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.
 One of the best features of PaRIS algorithm is that it may be implemented online, using the observations $(Y_k)_{k\ge 1}$ as they are received, without any increasing storage requirements.

@@

In this paper, we propose to use SMC methods to obtain asymptotically unbiased approximations of smoothing expectations of POD processes by extending the PaRIS algorithm.
The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities.
 The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of the PaRIS algorithm to POD processes is that the acceptance rejection mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates. 
 The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights. 
 The linear version of the FFBSi algorithm proposed in \cite{douc:garivier:moulines:olsson:2011} could be extended in a similar way as PaRIS algorithm but it would still require a backward pass and would not be an online smoother. 
 The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to POD processes but also to any general state space model where the transition density of the hidden chain may be approximated using a positive and unbiased estimator.

Section~\ref{sec:model} describes the model and the smoothing quantities to be estimated. Section~\ref{sec:rwparis} provides the algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states. This section also details the application of this algorithm when the transition density may be approximated using a GPE. 
In Section~\ref{sec:convergence}, classical convergence results for SMC smoothers are extended to the setting of this paper and illustrated with numerical experiments in Section~\ref{sec:exp}. 
All proofs are postponed to Appendix~\ref{sec:append:proofs}.

\section{Model and framework}
\label{sec:model}
%. These estimators are designed for stochastic processes
Let $(X_t)_{t\ge 0}$ be defined as a weak solution to the following SDE in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \Gamma(X_t)\rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion on $\mathbb{R}^d$, $\alpha: \mathbb{R}^d\to\mathbb{R}^d$ and $\Gamma: \mathbb{R}^d\to \mathbb{R}^{d\times d}$. The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$.  In the following, for all $0 \le k \le n$, the state $X_{t_k}$ at time $k$ is referred to as $X_k$. For all $0\le k \le n$, the distribution of $Y_k$ given $X_k$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot)$. For the sake of simplicity, we use the shorthand notation $g_k(X_k)$ for $g(X_k,Y_k)$. 
The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
For all $0\le k \le n-1$, the conditional distribution of $X_{k+1} $ given $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.

@@

The algorithm proposed in this paper is based on Sequential Monte Carlo methods which offer a flexible framework to approximate such distributions with weighted empirical measures associated with random samples. 
At each time step, the samples are moved randomly in $\mathbb{R}^d$ and  associated with importance weights. 
In general situations, the computation of these importance weights involve the unknown transition density of the process \eqref{eq:target:sde}. 
The solution introduced in Section~\ref{sec:rwparis} requires an unbiased estimator of these unknown transition densities. Moreover, this estimator must be almost surely positive and upper bounded.
Statistical inference of stochastic differential equations is an active area of research and several solutions have been proposed to design unbiased estimates of transition densities. 
Those estimators require different assumptions on the model \eqref{eq:target:sde}, we provide below several solutions that can be investigated.

@@

In addition to provide an unbiased estimate of the transition density, the GPE ensure that this estimate is almost surely positive. 
Moreover, as detailled below, one can define, under certain conditions, a GPE that is almost surely upper bounded.

@@

In the case the previous assumptions are not fulfilled, alternatives to GPEs are given by continuous importance sampling procedures for SDE. 

@@

This analytical expression of the transition density is not tractable and is estimated by updating random samples at random times using tractable proposal distributions (for instance based on an Euler discretization of the original SDE). 

@@

The random samples are weighted using the Kolmogorov forward operator associated with the SDE which relies on the first two order derivatives  of the drift and diffusion coefficients and are therefore tractable.

The unbiasedness of these procedures as long as the controls of the variability of the estimates require moments assumptions and Holder type conditions on the parameters of the SDE \eqref{eq:target:sde}. 
Their efficiency require a fair amount of tuning as they highly depend on the proposal densities used to obtain the Monte Carlo samples and the point processes generating the underlying random times.
In addition to unbiasedness, the proposed algorithm in this work requires that the estimator of the transition density is almost surely positive and upper bounded. 
It is beyond the scope of this paper to define a continuous importance sampling estimator with such properties, but this could be an interesting perspective if it requires less assumptions than the GPE.
\section{The Generalized Random PaRIS algorithm}
\label{sec:rwparis}
The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
\begin{equation}
\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
\end{equation}
The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions. 
Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.

@@

Consider the following assumption: for all  $0\leq k\leq n-1$, there exists a random variable $\hat{\sigma}^k_+$ measurable with respect to $\mathcal{G}_{k+1}^N$ such that,
\begin{assumpt}
\label{assumpt:global}
\mathrm{sup}_{x,y,\zeta}\;\widehat{q}_k(x,y;\zeta)\leq \hat{\sigma}^k_+\eqsp.
\end{assumpt}
\begin{lemma}
\label{lem:AR:unbiased}
Assume that \eqref{assumpt:global} holds for some $0\le k\le n-1$.  For all $1\le i \le N$, define the random variable $J_k^i$ as follows:
\begin{algorithmic}
\REPEAT
\STATE Sample independently $\zeta$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
\UNTIL{$U \leq \widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta)/\hat{\sigma}^k_+$.}
\STATE Set $J_k^i = J$.
%\STATE \textsc{Sampling step:} Sample independently $
%\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
%\IF{ $U \leq \widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)/\hat{\sigma}_+$,}
%\STATE Set $J_k^{i,\ell} = J$.
%\ELSE 
%\STATE Return to \textbf{\sc Sampling Step}
%\ENDIF
\end{algorithmic}
Then, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of $J_{k}^{i}$ is $\Lambda_{k}^N(i,\cdot)$. 
\end{lemma}
\begin{proof}
See Appendix \ref{sec:append:proofs}.
\end{proof}
Note that Lemma~\ref{lem:AR:unbiased} still holds if assumption \eqref{assumpt:global} is relaxed and replaced by one of the two following assumptions:
\begin{assumpt}\label{assumpt:ysupport}
\mathrm{sup}_{j,y,\zeta}\;\widehat{q}_k(\xi^j_k,y,\zeta)\leq \hat{\sigma}^k_+\;.
\end{assumpt}
\vspace{-\baselineskip}
\begin{assumpt}\label{assumpt:local}
\mathrm{sup}_{i,j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^i_{k+1},\zeta)\leq \hat{\sigma}^k_+\;.
\end{assumpt}
It is worth noting that under assumptions \eqref{assumpt:global} or \eqref{assumpt:ysupport}, the linear complexity property of PaRIS algorithm still holds, whereas if only assumption \eqref{assumpt:local} holds, the algorithm has a quadratic complexity.

>>>>

\begin{abstract}
This paper introduces a new algorithm to approximate smoothed additive functionals for partially observed diffusion processes. 
This method relies on a new Sequential Monte Carlo method which allows to compute such approximations online, i.e. as the observations are received, and with a computational complexity growing linearly with the number of Monte Carlo samples. The original algorithm cannot be used in the case of partially observed stochastic differential equations since the transition density of the latent data is usually unknown. 
We prove that it may be extended to partially observed continuous processes by replacing this unknown quantity by an unbiased estimator obtained for instance using general Poisson estimators. 
This estimator is proved to be consistent and its performance are illustrated using data from two models. 
\end{abstract}

@@

\section{Introduction}
This paper introduces a new algorithm to solve the smoothing problem for hidden Markov models (HMMs) whose hidden state is a solution to a stochastic differential equation (SDE). These models are referred to as partially observed diffusion (POD) processes in \cite{olsson:strojby:2011}. The hidden state process $(X_t)_{t\ge 0}$ is assumed to be a solution to a SDE and the only information available is given by noisy observations $(Y_{k})_{0\le k\le n}$ of the states $(X_k)_{0\le k\le  n}$ (where $X_k$ stands for $X_{t_k}$) at some discrete time points $(t_k)_{0\le k\le n}$. The bivariate stochastic process $\{(X_{k},Y_{k})\}_{0\le k\le n}$ is a state space model such that conditional on the state sequence $(X_{k})_{0\le k\le n}$ the observations $(Y_{k})_{0\le k\le n}$ are independent and for all $0\le \ell\le n$ the conditional distribution of $Y_{\ell}$ given $\{X_{k}\}_{0\le k\le n}$ depends on $X_{\ell}$ only.

@@

However, in most cases the exact computation of these expectations is usually not possible explicitly. Sequential Monte Carlo (SMC) methods are popular algorithms to approximate smoothing distributions with random particles associated with importance weights. 
 \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with resampling steps to duplicate or discard particles according to their importance weights.
In the case of HMMs, approximations of the smoothing distributions may be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. 
Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. 
Then, a backward pass is performed to compute new weights (FFBS) or sample trajectories (FFBSi) in order to approximate the smoothing distributions. Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.\ using the observations as they are received) smoothed expectations of additive functionals. 
Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.
 One of the best features of PaRIS algorithm is that it may be implemented online, using the observations $(Y_k)_{k\ge 0}$ as they are received, without any increasing storage requirements.

@@

In this paper, we propose to use SMC methods to obtain consistent approximations of smoothing expectations of POD processes by extending the PaRIS algorithm.
The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities.
 The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of the PaRIS algorithm to POD processes is that the acceptance rejection mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates. 
 The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights. 
 The linear version of the FFBSi algorithm proposed in \cite{douc:garivier:moulines:olsson:2011} could be extended in a similar way as PaRIS algorithm but it would still require a backward pass and would not be an online smoother. 
 The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to POD processes but also to any general state space model where the transition density of the hidden chain may be approximated using a positive and unbiased estimator.

Section~\ref{sec:model} describes the model and the smoothing quantities to be estimated. Section~\ref{sec:rwparis} provides the algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states. This section also details the application of this algorithm when the transition density are approximated using a GPE. 
In Section~\ref{sec:convergence}, classical convergence results for SMC smoothers are extended to the setting of this paper and illustrated with numerical experiments in Section~\ref{sec:exp}. 
All proofs are postponed to Appendix~\ref{sec:append:proofs}.

\section{Model and framework}
\label{sec:model}
%. These estimators are designed for stochastic processes
Let $(X_t)_{t\ge 0}$ be defined as a weak solution to the following SDE in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \Gamma(X_t)\rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion on $\mathbb{R}^d$, $\alpha: \mathbb{R}^d\to\mathbb{R}^d$ and $\Gamma: \mathbb{R}^d\to \mathbb{R}^{d\times d}$. The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$.  In the following, for all $0 \le k \le n$, the state $X_{t_k}$ at time $k$ is referred to as $X_k$. For all $0\le k \le n$, the distribution of $Y_k$ given $X_k$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot)$. For the sake of simplicity, the shorthand notation $g_k(X_k)$ for $g(X_k,Y_k)$ is used. 
The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
For all $0\le k \le n-1$, the conditional distribution of $X_{k+1} $ given $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.

@@

The algorithm proposed in this paper is based on Sequential Monte Carlo methods which offer a flexible framework to approximate such distributions with weighted empirical measures associated with random samples. 
At each time step, the samples are moved randomly in $\mathbb{R}^d$ and  associated with importance weights. 
In general situations, the computation of these importance weights involve the unknown transition density of the process \eqref{eq:target:sde}. 
The solution introduced in Section~\ref{sec:rwparis} requires an unbiased estimator of these unknown transition densities. Moreover, this estimator must be almost surely positive and upper bounded.
Statistical inference of stochastic differential equations is an active area of research and several solutions have been proposed to design unbiased estimates of these transition densities. 
Those estimators require different assumptions on the model \eqref{eq:target:sde}, we provide below several solutions that can be investigated.

@@

In addition to provide an unbiased estimate of the transition density, the GPE ensure that this estimate is almost surely positive.  Moreover, as detailed below, under additional conditions, a GPE that is almost surely upper bounded can be defined.

@@

In the case the previous assumptions are not fulfilled, in particular assumption i),  alternatives to GPEs are given by continuous importance sampling procedures for SDE. 

@@

This analytical expression of the transition density is not tractable and is estimated by updating random samples at random times between $t_k$ and $t_{k+1}$ using tractable proposal distributions (for instance based on an Euler discretization of the original SDE). 

@@

The random samples are weighted using the Kolmogorov forward operator associated with the SDE which relies on the first two order derivatives  of the drift and diffusion coefficients (and is therefore tractable).

The unbiasedness of these procedures and  the controls of the variability of the estimates require moments assumptions and Holder type conditions on the parameters of the SDE \eqref{eq:target:sde}. 
Their efficiency require a fair amount of tuning as they highly depend on the proposal densities used to obtain the Monte Carlo samples and the point processes generating the underlying random times.
In addition to unbiasedness, the proposed algorithm in this work requires that the estimator of the transition density is almost surely positive and upper bounded. This implies additional assumptions on the SDE depending on the chosen estimate and could lead to interesting perspectives.

\section{The Generalized Random PaRIS algorithm}
\label{sec:rwparis}
The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
\begin{equation}
\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
\end{equation}
The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions. 
Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.

@@

Consider the following assumption: for all  $0\leq k\leq n-1$, there exists a random variable $\hat{\sigma}^k_+$ measurable with respect to $\mathcal{G}_{k+1}^N$ such that,
\begin{assumpt}
\label{assumpt:global}
\mathrm{sup}_{x,y,\zeta}\;\widehat{q}_k(x,y;\zeta)\leq \hat{\sigma}^k_+\eqsp.
\end{assumpt}
\begin{lemma}
\label{lem:AR:unbiased}
Assume that \eqref{assumpt:global} holds for some $0\le k\le n-1$.  For all $1\le i \le N$, define the random variable $J_k^i$ as follows:
\begin{algorithmic}
\REPEAT
\STATE Sample independently $\zeta$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
\UNTIL{$U \leq \widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta)/\hat{\sigma}^k_+$.}
\STATE Set $J_k^i = J$.
%\STATE \textsc{Sampling step:} Sample independently $
%\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
%\IF{ $U \leq \widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)/\hat{\sigma}_+$,}
%\STATE Set $J_k^{i,\ell} = J$.
%\ELSE 
%\STATE Return to \textbf{\sc Sampling Step}
%\ENDIF
\end{algorithmic}
Then, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of $J_{k}^{i}$ is $\Lambda_{k}^N(i,\cdot)$. 
\end{lemma}
\begin{proof}
See Appendix \ref{sec:append:proofs}.
\end{proof}
Note that Lemma~\ref{lem:AR:unbiased} still holds if assumption \eqref{assumpt:global} is relaxed and replaced by:
\begin{assumpt}\label{assumpt:ysupport}
\mathrm{sup}_{j,y,\zeta}\;\widehat{q}_k(\xi^j_k,y,\zeta)\leq \hat{\sigma}^k_+\;.
\end{assumpt}
It is worth noting that under assumptions \eqref{assumpt:global} or \eqref{assumpt:ysupport}, the linear complexity property of PaRIS algorithm is ensured. The following assumption can also be considered. For all $1\le i \le N$,
\begin{assumpt}\label{assumpt:local}
\mathrm{sup}_{j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^i_{k+1},\zeta)\leq \hat{\sigma}^{k,i}_+\;.
\end{assumpt}
If only assumption \eqref{assumpt:local} holds, the algorithm has a quadratic complexity. The bound of \eqref{assumpt:global} is uniform (it does not depend on the particles) and can be used for every particle $1\le i\le N$.  However, this  bound  can be large (with respect to the simulated set of particles) for the algorithm of Lemma~\ref{lem:AR:unbiased}.
The bound of \eqref{assumpt:local} requires $N$ computations per particle (therefore $N^2$ computations). However, it is clear that this second bound is sharper that the one of  \eqref{assumpt:global} for the acceptance rejection procedure and may lead to a computationally more efficient algorithm.

\paragraph{Bounded estimator of $q_k$ using GPEs}
For $x, y \in \mathbb{R}^d$, by Girsanov and Ito's formulas, the transition density $q_k(x,y)$ of \eqref{eq:target:sde} satisfies, with $\Delta_k = t_{k+1}-t_k$,
\begin{align*}
q_k(x,y)=\varphi_{\Delta_k}(x,y)\exp\left\lbrace A(y)-A(x)\right\rbrace \mathbb{E}_{\mathbb{W}^{x,y,\Delta_k}}\left[ \exp \left\lbrace - \int_0^{\Delta_k} \phi(\mw_s)\rmd s \right\rbrace \right]\eqsp,
\end{align*}
where $\mathbb{W}^{x,y,\Delta_k}$ is the law of Brownian bridge starting at $x$ at 0 and hitting $y$ at $\Delta_k$, $(\mw_t)_{0\leq t \leq \Delta_k}$ is such a Brownian bridge, $\varphi_{\Delta_k}(x,y)$ is the p.d.f. of a normal distribution with mean $x$ and variance $\Delta_k$, evaluated at $y$ and $\phi:\mathbb{R}^d\to\mathbb{R}$ is defined as:
\[
\phi(x) =\left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2\eqsp.
\]
Assume that there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta_k$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$. The performance of the estimator depends on the choice of $\Lo_\mw$ and $\U_\mw$ which is specific to the SDE. In the case of the models analyzed in Section~\ref{sec:exp} these bounds are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008} for the SINE model and in \cite{olsson:strojby:2011} for the log-growth model. Note that in the case where $\phi$ is not upper bounded, \cite{beskos:papaspiliopoulos:roberts:2008} proposed the EA3 algorithm. This {\em layered Brownian bridge} construction first samples random variables to determine in which {\em layer} the Brownian bridge lies before simulating the bridge conditional on the event that it belongs to the layer. By continuity of $\phi$, $\Lo_\mw$ and  $\U_\mw$ can be computed easily.
